---
title: "Appendix C: Data Extraction and Wrangling"
author: "Christina Wang and Fei Wang"
date: "December 13, 2015"
output: pdf_document
---

```{r setup, include=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(mosaic)
require(readr)
require(tidyr)
```

## Download Data

This appendix describes the broader dataset and details about how to create the analytic dataset used here. The complete dataset has a size of 170.7 MB and is available as `CollegeScorecard_Raw_Data.zip` at https://collegescorecard.ed.gov/data/. The website also contains a number of `Featured Downloads` that provide quick access to some of the data in which users may be most interested. 

The package, after unzipping, includes 18 comma-separated value (csv) files of annually compiled data, ranging from 1996 to 2013 (e.g. data from year 1996 is stored in file `MERGED1996_PP.csv`), a data dictionary, and other documentations. Please note that variabels and descriptions may change in different years. 

For our analysis, we use dataset `MERGED2011_PP.csv`. Consult data dictionary
`CollegeScorecardDataDictionary
-09-12-2015.pdf` for labels and notes on each variable contained in the dataset. 


## Import data

Copy the unzipped file `MERGED2011_PP.csv` (size: 119.3MB) into the folder where you saved your R Markdown file. The size of the file is 365KB. Set the working directory to your current folder. Run the following code (or R markdown file) in the same folder where you have copied the file. You may need to install `readr` package to run the `read_csv` function.  

```{r, warning=FALSE}
df2011 <- read_csv("MERGED2011_PP.csv")
```

## Select variables based on codebook

The full explanation for the names of the variables can be found in Codebook. 

```{r}
df_full <- df2011 %>%
  filter(HIGHDEG >= 3) %>% #choose institutions offering Graduate and Bachelor degree
  select(INSTNM, ZIP, st_fips, region, CONTROL, # UNITID for some reason cannot be selected
         DEBT_MDN_SUPP, GRAD_DEBT_MDN, GRAD_DEBT_N, WDRAW_DEBT_MDN, WDRAW_DEBT_N,
         mn_earn_wne_p10, md_earn_wne_p10, pct10_earn_wne_p10,
         pct25_earn_wne_p10, pct75_earn_wne_p10, pct90_earn_wne_p10, 
         ADM_RATE, PCTPELL, C150_4)

df_full <- df_full %>%
  mutate(DEBT_MDN_SUPP = extract_numeric(DEBT_MDN_SUPP),
         GRAD_DEBT_MDN = extract_numeric(GRAD_DEBT_MDN),
         GRAD_DEBT_N = extract_numeric(GRAD_DEBT_N),
         WDRAW_DEBT_MDN = extract_numeric(WDRAW_DEBT_MDN),
         WDRAW_DEBT_N = extract_numeric(WDRAW_DEBT_N),
         mn_earn_wne_p10 = extract_numeric(mn_earn_wne_p10),
         md_earn_wne_p10 = extract_numeric(md_earn_wne_p10),
         pct10_earn_wne_p10 = extract_numeric(pct10_earn_wne_p10),
         pct25_earn_wne_p10 = extract_numeric(pct25_earn_wne_p10),
         pct75_earn_wne_p10 = extract_numeric(pct75_earn_wne_p10),
         pct90_earn_wne_p10 = extract_numeric(pct90_earn_wne_p10),
         ADM_RATE = extract_numeric(ADM_RATE),
         PCTPELL = extract_numeric(PCTPELL),
         C150_4 = extract_numeric(C150_4)
  )
```

## Save the dataset 

The filtered dataset based on the raw dataset of MERGED2011_PP.csv. is stored as `df2011.rda`.

```{r}
save(df_full, file = "df2011.rda")
write.csv(df_full, file = "df2011.csv")
# The size of `df2011.rda` is 127.7KB
```



