---
title: "Observations needed in Constructing a Confidence Interval"
author: "Fei Wang, STAT495"
date: "September 28, 2015"
output:
  slidy_presentation: default
  beamer_presentation:
    colortheme: dolphin
    fonttheme: structurebold
    theme: AnnArbor
---
```{r message=FALSE, echo=FALSE}
require(mosaic)
options(digits=3)
trellis.par.set(theme=theme.mosaic())
```

# Introduction

## Statistical Rules of Thumb 1.14
- Precision does not vary linearly with sample size
- Width of CI is inversely proportional to the square root of the number of observations
- Sample size, an estimate of variability, pivotal variable (t-statistics in normal distribution)

# Assessment

How many observations do you think are needed for constructing a confidence interval?

# Rule of thumb

The width of a confidence interval, involving estimation of variability and sample size, decreases rapidly until 12 observations and then decreases less rapidly.

# Illustration

```{r, fig.keep='last',echo=FALSE}
A <- c()
for(i in 1:30){
  A[i] <- ((50+1.96*(10/sqrt(i)))-(50-1.96*(10/sqrt(i))))
}

plot(A, type = "l", main = "Relationship between Width of 95% CI and Sample Size", 
    xlab = "Sample Size", ylab = "Width of CI")
abline(v=12, col = "red")
```

# Simulation
```{r, fig.keep='last', echo=FALSE}
pop <- rnorm(100, mean = 50, sd = 10)


size <- c()
for(i in 1:30){
  size[i] <- i}

width <- c()
a <- function(){
  Width[i] <- function(){
  sample <- sample(pop, size = size, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(size))-(mean-1.96*sd/sqrt(size))
  }
  a[i] <- c(sapply(1:100, Width[i]))
  }








Width2 <- function(i){
  sample <- sample(pop, size = 2, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(2))-(mean-1.96*sd/sqrt(2))
}
a2 <- c(sapply(1:100, Width2))
densityplot(a2)

Width3 <- function(i){
  sample <- sample(pop, size = 3, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(3))-(mean-1.96*sd/sqrt(3))
}
a3 <- c(sapply(1:100, Width3))

Width4 <- function(i){
  sample <- sample(pop, size = 4, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(4))-(mean-1.96*sd/sqrt(4))
}
a4 <- c(sapply(1:100, Width4))

Width5 <- function(i){
  sample <- sample(pop, size = 5, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(5))-(mean-1.96*sd/sqrt(5))
}
a5 <- c(sapply(1:100, Width5))

Width8 <- function(i){
  sample <- sample(pop, size = 8, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(8))-(mean-1.96*sd/sqrt(8))
}
a8 <- c(sapply(1:100, Width8))

Width10 <- function(i){
  sample <- sample(pop, size = 10, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(10))-(mean-1.96*sd/sqrt(10))
}
a10 <- c(sapply(1:100, Width10))

Width11 <- function(i){
  sample <- sample(pop, size = 11, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(11))-(mean-1.96*sd/sqrt(11))
}
a11 <- c(sapply(1:100, Width11))

Width12 <- function(i){
  sample <- sample(pop, size = 12, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(12))-(mean-1.96*sd/sqrt(12))
}
a12 <- c(sapply(1:100, Width12))

Width13 <- function(i){
  sample <- sample(pop, size = 13, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(13))-(mean-1.96*sd/sqrt(13))
}
a13 <- c(sapply(1:100, Width13))

Width14 <- function(i){
  sample <- sample(pop, size = 14, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(14))-(mean-1.96*sd/sqrt(14))
}
a14 <- c(sapply(1:100, Width14))

Width15 <- function(i){
  sample <- sample(pop, size = 15, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(15))-(mean-1.96*sd/sqrt(15))
}
a15 <- c(sapply(1:100, Width15))

Width18 <- function(i){
  sample <- sample(pop, size = 18, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(18))-(mean-1.96*sd/sqrt(18))
}
a18 <- c(sapply(1:100, Width18))

Width20 <- function(i){
  sample <- sample(pop, size = 20, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(20))-(mean-1.96*sd/sqrt(20))
}
a20 <- c(sapply(1:100, Width20))

Width30 <- function(i){
  sample <- sample(pop, size = 30, replace = TRUE)
  mean <- mean(sample)
  sd <- sd(sample) 
  width <- (mean+1.96*sd/sqrt(30))-(mean-1.96*sd/sqrt(30))
}
a30 <- c(sapply(1:100, Width30))

plot(density(a2), xlim=c(0,40), ylim=c(0,0.3), 
    main= "Distribution of Width of CI for Different Sample Size", xlab="Width of CI", ylab="Density") 
lines (density(a3))
lines (density(a4))
lines (density(a5))
lines (density(a8))
lines (density(a10),col="blue")
lines (density(a11),col="green")
lines (density(a12),col="red")
lines (density(a13),col="yellow")
lines (density(a14),col="pink")
lines (density(a15))
lines (density(a18))
lines (density(a20))
```

Theoretical: 27.72 22.63 19.60 17.53 13.86 12.40 11.82 11.32 10.87 10.48 10.12 9.24  8.77  
Blue:10, Green:11, Red:12, Yellow:13, Pink:14

-----


## Closing thoughts

- The width of CI decreases as sample size increases
- If variance has to be estimated from the data, uncertainty due to this estimation, which is incorporated in the t-statistics, decreases also as the sample size increases


